---
title: "Basic Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic_usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(teaR)
library(phyloseq)
```

First, let's load the data  
```{r load_data}
data(GlobalPatterns)
```

We can use many ways to construct sets. The easiest way is to construct sets based on taxonomic ranks. Then we can attach these sets to a `phyloseq` object.  
```{r const_set_rank}
family_sets <- const_set(tax_table(GlobalPatterns), rank = "Family")
physeq <- phyloseqSet(GlobalPatterns, taxon_set = family_sets)
print(physeq)
```
This shows that among 19216 taxa that are on hand, about 13613 taxa are part of 341 sets. The number of taxa in taxon set tables needs to be a subset of all taxa but can be smaller than the total number of taxa. This is because we can filter sets based on a variety of criteria and therefore would only use the subset of total taxa that are assigned to those sets. To filter sets, we can use the `trim_sets()` function, which can take multiple criteria as long as they have been attached to the `BiocSet` object itself. The constructor function automatically adds a `size` column to the `set` slot in the `BiocSet` object, which means that we can always filter by size. 

```{r trim_set}
physeq <- trim_set(physeq, size >= 5)
physeq
```

The reason why `phyloseqSet` was used as the extension is that we can still apply `phyloseq` operations and it would affect the resulting `taxon_set()` slot. This is because if we filtered out taxa due to issues like low abundance, then those taxa would also not appear in the elements section of taxon sets. However, **order matters**, especially if sets are being filtered based on size. If a `filter_taxa` step was performed after `taxon_set()` was attached, then the resulting sets need to be trimmed down further. The function `trim_sets()` can be used to filter taxa by different criteria.  


